\large{Curtin University} \\ \large{Curtin Institute of Radio Astronomy} \\ \large{Physics Project 1} \\ \vspace*{20px} \textbf{\LARGE{Finding New Pulsars Using Machine Learning}}\vspace*{10px}

### 2020-06-03 Draft.tex
methods

    \maketitle

@@

        As radio telescope technology improves, the number of candidate detections to be classified as pulsar or non-pulsar during pulsar surveys increases to the point that it becomes infeasible to classify them manually, a state of affairs known as the "pulsar candidate selection problem". To solve this problem, machine learning (ML) techniques are used to filter out the non-pulsar candidates in surveys, reducing the number of candidates requiring human evaluation to a manageable number. Here, we adapt an ML solution developed for the LOFAR Tied-Array All-Sky Survey (LOTAAS) for the Southern-sky Murchison Widefield Array Rapid Two-metre (SMART) survey. A group of 147 pulsar and 87 non-pulsar candidates generated by the SMART survey were used to test the LOTAAS ML classifiers trained on a set of 12 pulsar and 13 non-pulsar SMART survey candidates. The classifiers were found, on average, to make accurate pulsar and non-pulsar classifications on SMART survey candidates 87.3\% of the time, compared to the 97.1\% success rate obtained by \textcite{lyon}. Using the machine learning classifiers in ensemble with a selection criteria of three separate positive classifications increased the success rate of non-pulsar classification by 5.2\% (making 0 False Positive classifications), however it decreased the success rate of pulsar classification by 3.5\%. The ensemble machine learning classifier can be further improved by increasing the diversity of pulsars within the machine learning training dataset.

@@

To maximise the accuracy of the machine learning algorithm, the input data (including the training dataset) must be composed of a common group of features that can be determined for each candidate that maximises the differences between a pulsar and a non-pulsar. The candidate features used by \smartcite{tan} to maximise the differences between pulsar and non-pulsar candidates are:

@@

Where candidate features are calculated from the: (1) Integrated Pulsar Profile, (2) the Dispersion Measure -- Signal-to-Noise Ratio Curve (DM-S/N), (3) the correlation coefficients between each sub-band and the integrated pulsar profile, and from the (4) correlation coefficients between each sub-integration and the integrated pulsar profile.

To extract the 20 above features from each classification candidate, we can use the software 
\verb|PulsarFeatureLab| \autocite{lyon}.

\subsubsection{Pulsar Candidate Feature Extraction}

The Python software tool \verb|PulsarFeatureLab| can be used to process pulsar candidate files of the PRESTO Prepfold \verb|PFD| filetype and output the 20 above features for each candidate into a single file of WEKA Data Mining \verb|ARFF| filetype \autocite{lyon}.

To create a closed software environment in which the dependencies of the \verb|PulsarFeatureLab| software are unaffected by the host operating system, a containerised virtual operating system was created using the free software Docker (\verb|https://docker.com|).

Candidate \verb|PFD| files of known pulsars and non-pulsars detected by the Murchison Widefield Array (MWA) provided by N. Swainston can now populate the above created directory, and the following command can be ran to extract the features from the candidates:

This function instructs Docker to connect the directory containing the \verb|PFD| files to the \verb|PulsarFeatureLab| container's input/output directory and then run the \verb|PulsarFeatureLab| software with arguments stating where the input files are, what filetype they are (\verb|PFD|), which set of features to extract, and where to place the output file.

\subsubsection{Creating the Training Dataset}

The \verb|output.arff| file created by \verb|PulsarFeatureLab|, contains a set of comma-separated features for each candidate, with an appended `?' character, per file line. Since the class of each candidate is already known, the `?' character on each line can be replaced by a `1' if the candidate is a pulsar, or a `0' if the candidate is a non-pulsar. This signals to the Machine Learning algorithm what a pulsar and a non-pulsar candidate's feature set may appear like.

The edited file can now be renamed and moved with the following command:

\begin{lstlisting}[numbers=none]
$ mv ~/pulsars/data/pfd/output.arff ~/pulsars/data/trainingSet.arff
\end{lstlisting}

The Machine Learning Training Dataset has now been created using Murchison Widefield Array data.

\subsection{Evaluating the LOTAASClassifier Machine Learning Classification Tool for use with the Murchison Widefield Array}

@@

\subsubsection{Creating a Pulsar Classifier Output Validation Tool}
\label{sec:methodvalidate}
In order to accurately evaluate and test the predictions made by the \verb|LOTAASClassifier| tool, we must create a program that will automate the process of checking the outputs of the classifier against the known list of pulsars.

We begin by creating a new Java project using the free software Maven (\verb|https://maven.apache.org|) by running the following commands in a UNIX terminal:

This software will consume a user inputted \verb|String| containing the path to a file with the list of pulsars included in the dataset classified by the machine learning classifier, a user inputted \verb|String| containing the path to the \verb|.positive| file created by the classifier, and a user inputted \verb|String| containing the path to the \verb|.negative| file created by the classifier\footnote{The '.positive' file contains the candidates classified as pulsars, and the '.negative' file contains the candidates classified as non-pulsars.}. These will be inputted as command-line arguments when the user runs the Java executable file.

\subsubsection{Creating a Classification Model}

@@

    \section{Methods}
    \begin{subappendices}
        \subsection{Feature Extraction with PulsarFeatureLab}

        \subsection{Using PulsarValidator}

@@

    \pagebreak
    \section{Pulsar Validator}
    \begin{subappendices}
        \label{sec:pulsarvalidator} 
        \subsection{PulsarValidator.java}
        \lstinputlisting[title=PulsarValidator.java, language=Java, basicstyle=\footnotesize\ttfamily, breaklines=true, framexleftmargin=2.5em]{code/jacobianmatthews/pulsarvalidator/PulsarValidator.java}

@@

    \pagebreak
    \section{Pulsar Classifier}
    \begin{subappendices}
        \label{sec:pulsarclassifier}
        \subsection{PulsarClassifier.java}
        \lstinputlisting[title=PulsarClassifier.java, language=Java, basicstyle=\footnotesize\ttfamily, breaklines=true, framexleftmargin=2.5em]{code/jacobianmatthews/pulsarclassifier/PulsarClassifier.java}

>>>>

    \maketitle
    \thispagestyle{empty}

@@

        As radio telescope technology improves, the number of candidate detections to be classified as pulsar or non-pulsar during pulsar surveys increases to the point that it becomes infeasible to classify them manually, a state of affairs known as the ``pulsar candidate selection problem". To solve this problem, machine learning (ML) techniques are used to filter out the non-pulsar candidates in surveys, reducing the number of candidates requiring human evaluation to a manageable number. Here, we adapt an ML solution developed for the LOFAR Tied-Array All-Sky Survey (LOTAAS) for the Southern-sky Murchison Widefield Array Rapid Two-metre (SMART) survey. A group of 147 pulsar and 87 non-pulsar candidates generated by the SMART survey were used to test the LOTAAS ML classifiers trained on a set of 12 pulsar and 13 non-pulsar SMART survey candidates. The classifiers were found, on average, to make accurate pulsar and non-pulsar classifications on SMART survey candidates 87.3\% of the time, compared to the 97.1\% success rate obtained by \textcite{lyon}. Using the machine learning classifiers in ensemble with a selection criteria of three separate positive classifications increased the success rate of non-pulsar classification by 5.2\% (making 0 False Positive classifications), however it decreased the success rate of pulsar classification by 3.5\%. The ensemble machine learning classifier can be further improved by increasing the diversity of pulsars within the machine learning training dataset.

@@

To maximise the accuracy of the machine learning algorithm, the input data (including the training dataset) must be composed of a common group of features that can be determined for each candidate that maximises the differences between a pulsar and a non-pulsar. The candidate features used by \textcite{tan} to maximise the differences between pulsar and non-pulsar candidates are:

@@

Where candidate features are calculated from the: (1) Integrated Pulsar Profile, (2) the Dispersion Measure -- Signal-to-Noise Ratio Curve (DM-S/N curve), (3) the correlation coefficients between each sub-band and the integrated pulsar profile, and (4) from the correlation coefficients between each sub-integration and the integrated pulsar profile.

A set of 12 pulsar and 13 non-pulsar candidates from the SMART survey were provided by N. Swainston to become the machine learning training dataset. The free Python software created by \textcite{lyon}, \verb|PulsarFeatureLab|, was used to process the PRESTO Prepfold \verb|PFD| pulsar candidate files, extracting the above 20 machine learning features from each candidate into a single file of WEKA Data Mining \verb|ARFF| filetype. To create a closed software environment in which the dependencies of the \verb|PulsarFeatureLab| software were unaffected by the host operating system, a containerised virtual operating system was created using the free software Docker (\verb|https://docker.com|) and the pulsar feature extraction software was thus ran using a bind mounted volume. For further details on how this was achieved, see \emph{Appendix \ref{apndx:pulsarfeaturelab}}.

The file outputted by \verb|PulsarFeatureLab| containing the feature extracted candidates was then edited, removing the `?' character and appending a `1' or `0' to the end of each candidate's line depending on whether the candidate was a pulsar or non-pulsar respectively. This tells the machine learning classifier that candidates with features similar to those preceding the appended binary classification belong to a pulsar (or a non-pulsar). The remaining file constitutes the machine learning training dataset.

\subsection{Developing a Machine Learning Classifier Validation Tool}
\label{sec:methodvalidate}

In order to automate the evaluation of predictions made by the machine learning pulsar classifiers, a Java program named \verb|PulsarValidator| was developed.

\verb|PulsarValidator| is designed to load the positive and negative output files generated from the machine learning classification of a controlled testing dataset of pulsar candidates, and compare these files to a list of the pulsars included in the testing dataset. The software outputs the following statistics: the number of pulsars and non-pulsars, the number of detected pulsars and non-pulsars, and the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).

This was achieved by first creating individual Java `ArrayList's to contain the filenames of each candidate found to be a True Positive, False Positive, True Negative, or False Negative classification. The positive and negative output files generated by the classifier (positive denoting pulsars and negative denoting non-pulsars), and the list of pulsars in the testing dataset are then also converted to Java `ArrayList's. The items inside the positive and negative `ArrayList's are then iterated over, checking each item against the list of pulsars. If an item in the positive list is found in the list of pulsars, it is added to the True Positive list, otherwise it is added to the False Positive list. If an item in the negative list is found in the list of pulsars, it is added to the False Negative list, otherwise it is added to the True Negative list. The number of items inside each list is then outputted to the user. This process can be seen in \emph{Algorithm \ref{alg:validator}}. Further details on the development of this software can be found in \emph{Appendix \ref{apndx:pulsarvalidator}}, and the full Java source code can be found in \emph{Appendix \ref{src:pulsarvalidator}} or on GitHub (\verb|https://github.com/jacob-ian/PulsarValidator.git|).

\begin{algorithm}
    \caption{Pulsar Validator (pseudocode)}
    \label{alg:validator}
    \begin{algorithmic}
        \State Let $truePositive$ = new List
        \State Let $falsePositive$ = new List
        \State Let $trueNegative$ = new List
        \State Let $falseNegative$ = new List

\subsection{Evaluating the LOTAAS Machine Learning Classification Tool for use with the SMART Survey}

@@

@@

    \section{Methods}
    \begin{subappendices}
        \subsection{Feature Extraction with PulsarFeatureLab}
        \label{apndx:pulsarfeaturelab}

        For future development of this project, the Docker image \\\verb|jacobianm/pulsarfeaturelab:1.3.2| is available on Docker Hub\\(\verb|https://hub.docker.com|) and will be automatically downloaded when running the \verb|docker run| command below. Candidate \verb|PFD| files of known pulsars and non-pulsars detected by the Murchison Widefield Array (MWA) can now populate the above created directory, and the following command can be ran to extract the features from the candidates:

        This function instructs Docker to connect the directory containing the \verb|PFD| files to the \verb|PulsarFeatureLab| container's input/output directory and then run the \verb|PulsarFeatureLab| software with arguments stating where the input files are, what filetype they are (\verb|PFD|), which set of features to extract, and where to place the output file (\verb|output.arff|).

        \subsection{Developing PulsarValidator}
        \label{apndx:pulsarvalidator}
        We begin by creating a new Java project using the free software Maven (\verb|https://maven.apache.org|) by running the following commands in a UNIX terminal:

        This software will process a user inputted \verb|String| containing the path to a file with the list of pulsars included in the dataset classified by the machine learning classifier, a user inputted \verb|String| containing the path to the \verb|.positive| file created by the classifier, and a user inputted \verb|String| containing the path to the \verb|.negative| file created by the classifier\footnote{The '.positive' file contains the candidates classified as pulsars, and the '.negative' file contains the candidates classified as non-pulsars.}. These will be inputted as command-line arguments when the user runs the Java executable file.

        \subsection{Creating a Classification Model with LOTAASClassifier}

@@

    \pagebreak
    \section{Pulsar Validator}
    \begin{subappendices}
        \label{src:pulsarvalidator} 
        \subsection{PulsarValidator.java}
        \lstinputlisting[title=PulsarValidator.java, language=Java, basicstyle=\footnotesize\ttfamily, breaklines=true, framexleftmargin=2.5em]{code/jacobianmatthews/pulsarvalidator/PulsarValidator.java}

@@

    \pagebreak
    \section{Pulsar Classifier}
    \begin{subappendices}
        \label{src:pulsarclassifier}
        \subsection{PulsarClassifier.java}
        \lstinputlisting[title=PulsarClassifier.java, language=Java, basicstyle=\footnotesize\ttfamily, breaklines=true, framexleftmargin=2.5em]{code/jacobianmatthews/pulsarclassifier/PulsarClassifier.java}
